k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-02-29 paper.Rmd
abtract

---
title             : "An improved anticlustering method to maximize stimulus dispersion in experiments"
shorttitle        : "Maximum stimulus dispersion"

@@

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise within-set dissimilarity). We provide a first independent assessment of BILS and extend it in three ways. First, by prefixing an exact algorithm, our extension ensures that a globally optimal value of dispersion is always returned; without it, it is unknown if the BILS's result is optimal. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to prevent specific pairs of stimuli from appearing in the same condition. Third, our extension allows that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. A simulation study provides an extensive evaluation of the original BILS and our extension. Using the optimal approach as benchmark, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but we also highlight conditions where it fails to do so. A practical example shows how the extended BILS can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.

>>>>

---
title             : "An optimal anticlustering method to maximize stimulus dispersion in experiments"
shorttitle        : "Maximum stimulus dispersion"

@@

    When assigning stimuli to experimental conditions, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) to obtain a stimulus partitioning that simultaneously achieves high values according to two criteria: the diversity, corresponding to between-set similarity, and the dispersion, corresponding to pairwise within-set dissimilarity. As a heuristic, however, the BILS does not necessarily identify globally optimal criterion values. We present a novel exact algorithm that returns (multiple) globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, it scales to rather large data sets and can be applied to solve most real-world use cases. In a simulation study, we evaluate several alternatives to combine our optimal approach with the BILS heuristic, which -- due to its bicriterion approach -- naturally lends itself to providing high diversity on top of an optimal dispersion. We also provide two further extensions to the BILS: First, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices. Second, we allow that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. The original BILS as well as our extensions are freely available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.
